Analyzing Student Use of Spacing and Interleaving Strategies in Interactions with GenAI-Powered Chatbots in Programming Courses

### 2025-01-21 CSEDU2025/Paper.tex
paper update

\title{
  AAnalyzing Student Use of Spacing and Interleaving Strategies in Interactions
  with GenAI-Powered Chatbots in Programming Courses
}

@@

\abstract{
The aim of this study is to analyze the interactions of programming students
with a chatbot powered by OpenAI's GPT-3.5, enhanced with the Retrieval
Augmented Generation (RAG) technique, within the context of a Java programming
course. The study investigates how students use the chatbot while employing
two metacognitive strategies: interleaving and spacing. The methodology involved
classifying and categorizing the students' interaction prompts with the chatbot
into eight distinct categories. The findings reveal that, while most
interactions focus on active learning strategies, students demonstrate limited
conscious application of metacognitive strategies. The study highlights the need
for more comprehensive guidance on the strategic use of AI tools to enhance
learning outcomes.
}

@@

Spacing and interleaving are two metacognitive strategies that have been shown
to enhance learning outcomes and support the research questions of this study.
Spacing refers to the practice of distributing study sessions over time, which
has been shown to improve long-term retention and understanding of the material
\citep{Carvalho20}. Interleaving involves mixing different topics or problems
within a single learning session, which has been shown to also enhance long-term
learning and the application of student's knowledge in other contexts and
situations \citep{Rivers21}.

@@

\cite{Boudouaia24} investigated the impact of ChatGPT-facilitated programming on
college students’ programming behaviors, performance, and perceptions. The
problem addressed was whether the use of ChatGPT in programming education could
influence how students interacted with programming tasks, their learning
outcomes, and their attitudes towards technology. The study employed a
quasi-experimental design involving 82 college students divided into two groups:
one using ChatGPT-assisted Programming (CAP) and the other using traditional
self-directed Programming (SDP). The method combined mixed data collection
techniques, including behavioral logs, programming performance evaluations, and
interviews. The results showed that students using the CAP mode exhibited more
frequent behaviors such as debugging and reading feedback. Although CAP students
had a higher average score in programming performance, there was no
statistically significant difference compared to the SDP group. However,
students’ perceptions of ChatGPT improved significantly after the intervention,
with increased perceived usefulness, ease of use, and intention to continue
using the tool in the future.

\cite{Bastani24} investigated the impact of GenAI, specifically GPT-4, on human
learning, with a focus on mathematics education in a high school. The problem
addressed was how the use of generative AI could affect the acquisition of new
skills, which is crucial for long-term productivity. The method involved a
controlled experiment with approximately one thousand students, who were exposed
to two GPT-4-based tutors: a simple tutor (GPT Base) and another with safeguards
designed to promote learning (GPT Tutor). The results showed that while access
to GPT-4 improved performance on practice exercises (48\% with GPT Base and
127\% with GPT Tutor), the removal of access to GPT Base led to a 17\% decrease
in student performance on exams. This suggests that unrestricted use of
\textit{GPT Base} could hinder learning. However, the GPT Tutor was able to
mitigate this negative effect.
\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.1} % Increase the spacing between rows
    \begin{tabular}{p{3.8cm} p{6cm} p{4.5cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} & \textbf{Example of prompt} \\
      \hline
      Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. & \textit{Would this code be ok? \{code\}} \\
      Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. & \textit{What does it mean for a method to be static?} \\
      Student Correction (SC) & Prompts where the student corrects the chatbot. & \textit{The correct answer is B} \\
      Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. & \textit{A class inherits from another write this code} \\
      Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. & \textit{Give me the code for a selection sort} \\
      Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). & \textit{A heap is a data structure appropriate for: \{options\}} \\
      Language Change (LC) & Prompts that request a change of idiom. & \textit{In Spanish} \\
      Uncategorized (U) & Prompts that do not fit into any of the above categories. & \textit{Thanks} \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing
improvements in problem-solving and collaboration. \cite{Khusnul24} investigates
the efficacy of a meta-learning approach in improving metacognitive and creative
skills. This study confirms the effectiveness of metacognition strategies and
elucidates the relationship between meta-learning and metacognition. This
research suggests that meta-learning can improve metacognitive abilities,
providing valuable insights into educational technology and course design in
higher education settings.

\cite{Sidra24} addressed the growing importance of metacognitive skills in an
AI-enabled workforce. The problem discussed was how AI systems, especially in
collaborative tools based on Large Language Models (LLMs), impacted human
decision-making and the challenges they posed due to the lack of human-like
metacognition in AI. The method used identified four main characteristics that
differentiated human-AI interaction from human-human collaboration and proposed,
based on the dual-process theory of human thought, that metacognitive thinking
was crucial for the effective use of AI. The results indicated that human
workers needed to develop and apply metacognitive skills, such as planning,
monitoring, and evaluating their thought processes, to mitigate cognitive biases
and improve decision-making when working with AI. The study concluded by
recommending improvements in AI design and training to improve workers'
metacognitive abilities, ensuring more effective collaboration with AI systems.

@@

CharlieBot is an educational chatbot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) technique to improve the precision of
responses provided to students \citep{Sun24}. The RAG approach was developed
using materials from a MOOC, including text, exercises, video transcripts, and
other resources.

@@

The study comprised three phases: data collection, categorization, and analysis.
During the data collection phase, students enrolled in a second-semester Java
course at a public university in Spain were introduced to CharlieBot and given
the freedom to use it without following any prescribed educational methodology.
All data were collected anonymously to ensure that interactions could not be
traced back to individual students or linked to their academic performance.

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis, representing different types of
prompt that students use when interacting with the chatbot. Initially,
\cite{Ghimire24} proposed four categories: Debugging Help (DH), Conceptual
Question (CQ), Code Snippet (CS), and Complete Solution (CSO). However, the data
collected indicated the need for additional categories, leading to the inclusion
of four more: Multiple Question (MQ), Student Corrections (SC), Language Change
(LC), and Uncategorized (U). In this study, students were free to use the
chatbot as they wanted, which contributed to the emergence of these additional
categories. Table \ref{tab:categories} presents these categories along with
their respective descriptions. To facilitate the analysis, the prompt topics
were also categorized according to the course subjects, which are: (1) Java, (2)
Object Orientation, (3) Testing, (4) Recursion, (5) Data Structures, and (6)
Sorting and Searching Algorithms.

The analysis phase involved the use of Python scripts to extract information
from previously classified data. The data sets and code used in this process are
available at: https://github.com/rodrigoprestesmachado/uc3m.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.62]{img/figure1.png}
  \caption{Categorization of messages.}
  \label{fig:graph1}
\end{figure}

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.42]{img/figure2.png}
  \caption{Examples of conversations.}
  \label{fig:graph2}
\end{figure*}

@@

\begin{table*}[htbp]
  \caption{Average and Standard Deviation of Different Categories (in minutes)}
  \begin{center}
    \renewcommand{\arraystretch}{1.2} % Adjust the spacing between rows
    \begin{tabular}{p{4cm} p{3cm} p{3cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
      \hline
      Code Snippet & 1.34 & 1.95 \\
      Uncategorized & 1.70 & 3.59 \\
      Language Change & 2.23 & 3.71 \\
      Multiple Question & 2.47 & 5.66 \\
      \hdashline
      \textit{Overall} & \textit{3.78} & \textit{6.95} \\
      \hdashline
      Student Correction & 4.07 & 8.42 \\
      Debugging Help & 5.30 & 10.80 \\
      Conceptual Questions & 5.47 & 10.46 \\
      Complete Solution & 7.65 & 11.03 \\
      \hline
    \end{tabular}
    \label{tab:averages}
  \end{center}
\end{table*}

@@

Conversations without spacing comprise around 66\% of the total, with 34.2\%
showing a topic change, indicating interleaving (equivalent to 22.8\% of all
conversations). Qualitative analysis of these conversations without spacing
revealed that about 7\% of the conversations presented interleaving caused by
students' at least one request to solve quizzes. However, the speed with which
students interact with the chatbot to solve quizzes may suggest a more
superficial reflection on the answers received (see section 4.2.1). On the other
hand, 8.8\% of the conversations showed no spacing but included topic changes,
suggesting that students effectively might employ a metacognitive interleaving
strategy to distinguish concepts around a problem. Finally, 7\% of the
conversations featured interaction pauses lasting between 20 and 59 minutes but
fell short of the study's definition of spacing, which required pauses of over
60 minutes. This subset of conversations revealed a natural interleaving of
topics, reflecting a chatbot usage pattern that often mirrors how users search
Web pages.

@@

The results indicated that messages from the Code Snippets and Multiple Question
(Quiz) category implied quick and consistent interactions with the bot. This
pattern reflects a good level of engagement with the tool but also raises the
possibility of a lack of more profound reflection on the received content, given
that students' subsequent messages were sent consistently within short intervals.

The results also indicated that students' use of study strategies such
as Spacing and Interleaving remains limited. Therefore, encouraging students to
apply these strategies intentionally could enhance their learning using GenAI.

@@

There are ways to deepen the study; based on previous references, 60 minutes was
the standard spacing measure. However, further studies will deepen this analysis
using a quantitative variable to represent different spacing intervals better.
Future research can also include additional metacognitive strategies, which will
involve analyzing and interpreting user messages in depth. For instance, the
Monitoring strategy was evident when students expressed that a concept remained
unclear and requested further clarification. Another necessary analysis involves
investigating the relationship between student profiles, metacognitive
strategies for using GenAI, and learning outcomes.

>>>>

\title{
  Analyzing Student Use of Spacing and Interleaving Strategies in Interactions
  with GenAI-Powered Chatbots in Programming Courses
}

@@

\abstract{
This study aimed to analyze prompts of programming students with a chatbot
powered by OpenAI's GPT-3.5, enhanced with the Retrieval Augmented Generation
(RAG) technique, within the context of a Java programming course. The focus was
on students using two metacognitive strategies: interleaving and spacing.
Student prompts were categorized into eight categories along with their
respective study topics. Findings revealed that the categories and markers of
spacing and interleaving were important in identifying study sessions with the
chatbot. However, students showed limited intentional application of these
learning strategies. These results highlight the need for more comprehensive
guidance on leveraging AI tools to improve learning outcomes.
}

@@

Spacing and interleaving are two metacognitive strategies that have been shown
to enhance learning outcomes and support the research questions of this study.
Spacing refers to the practice of distributing study sessions over time, which
has been shown to improve long-term retention and understanding of the material
\citep{Carvalho20}. Interleaving involves mixing different topics or problems
within a single learning session, which has been shown to also enhance long-term
learning and the application of student's knowledge in other contexts and
situations \citep{rohrer20} \citep{Rivers21} .

@@

% \cite{Boudouaia24} investigated the impact of ChatGPT-facilitated programming on
% college students’ programming behaviors, performance, and perceptions. The
% problem addressed was whether the use of ChatGPT in programming education could
% influence how students interacted with programming tasks, their learning
% outcomes, and their attitudes towards technology. The study employed a
% quasi-experimental design involving 82 college students divided into two groups:
% one using ChatGPT-assisted Programming (CAP) and the other using traditional
% self-directed Programming (SDP). The method combined mixed data collection
% techniques, including behavioral logs, programming performance evaluations, and
% interviews. The results showed that students using the CAP mode exhibited more
% frequent behaviors such as debugging and reading feedback. Although CAP students
% had a higher average score in programming performance, there was no
% statistically significant difference compared to the SDP group. However,
% students’ perceptions of ChatGPT improved significantly after the intervention,
% with increased perceived usefulness, ease of use, and intention to continue
% using the tool in the future.

The study of \cite{Boudouaia24} examined the effects of ChatGPT-assisted
programming on university students' behaviors, performance, and perceptions.
A quasi-experimental research was conducted with 82 students divided into two
groups: one with ChatGPT-assisted programming (CAP) and the other with
self-directed programming (SDP). The analysis included behavioral logs,
performance evaluations, and interviews. Students in the CAP group engaged more
actively in debugging and feedback review activities. Although they achieved
slightly higher scores, there was no statistically significant difference in
performance compared to the SDP group. Nevertheless, perceptions of ChatGPT
improved significantly, highlighting greater perceived usefulness, ease of use,
and intention to use the tool in the future.

\cite{Bastani24} investigated the impact of GenAI, specifically GPT-4, on human
learning, with a focus on mathematics education in a high school. The problem
addressed was how the use of generative AI could affect the acquisition of new
skills, which is crucial for long-term productivity. The method involved a
controlled experiment with approximately one thousand students, who were exposed
to two GPT-4-based tutors: a simple tutor (GPT Base) and another with safeguards
designed to promote learning (GPT Tutor). The results showed that while access
to GPT-4 improved performance on practice exercises (48\% with GPT Base and
127\% with GPT Tutor), the removal of access to GPT Base led to a 17\% decrease
in student performance on exams. This suggests that unrestricted use of
\textit{GPT Base} could hinder learning. However, the GPT Tutor was able to
mitigate this negative effect.

\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.1} % Increase the spacing between rows
    \begin{tabular}{p{3.8cm} p{6cm} p{4.5cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} & \textbf{Example of prompt} \\
      \hline
      Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. & \textit{Would this code be ok? \{code\}} \\
      Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. & \textit{What does it mean for a method to be static?} \\
      Student Correction (SC) & Prompts where the student corrects the chatbot. & \textit{The correct answer is B} \\
      Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. & \textit{A class inherits from another write this code} \\
      Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. & \textit{Give me the code for a selection sort} \\
      Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). & \textit{A heap is a data structure appropriate for: \{options\}} \\
      Language Change (LC) & Prompts that request a change of idiom. & \textit{In Spanish} \\
      Uncategorized (U) & Prompts that do not fit into any of the above categories. & \textit{Thanks} \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing

% \cite{Khusnul24} investigates
% the efficacy of a meta-learning approach in improving metacognitive and creative
% skills. This study confirms the effectiveness of metacognition strategies and
% elucidates the relationship between meta-learning and metacognition. This
% research suggests that meta-learning can improve metacognitive abilities,
% providing valuable insights into educational technology and course design in
% higher education settings.

% \cite{Sidra24} addressed the growing importance of metacognitive skills in an
% AI-enabled workforce. The problem discussed was how AI systems, especially in
% collaborative tools based on Large Language Models (LLMs), impacted human
% decision-making and the challenges they posed due to the lack of human-like
% metacognition in AI. The method used identified four main characteristics that
% differentiated human-AI interaction from human-human collaboration and proposed,
% based on the dual-process theory of human thought, that metacognitive thinking
% was crucial for the effective use of AI. The results indicated that human
% workers needed to develop and apply metacognitive skills, such as planning,
% monitoring, and evaluating their thought processes, to mitigate cognitive biases
% and improve decision-making when working with AI. The study concluded by
% recommending improvements in AI design and training to improve workers'
% metacognitive abilities, ensuring more effective collaboration with AI systems.

\cite{Sidra24} explored the importance of metacognitive skills in an AI-enabled
workforce, focusing on how AI, particularly tools using Large Language Models,
affects human decision-making. They highlighted challenges arising from AI's
lack of human-like metacognition and identified key differences between human
AI and human-human collaboration. Using dual-process theory, they emphasized the
need for workers to develop metacognitive skills like planning, monitoring, and
evaluating to counter cognitive biases and enhance decision-making. The study
recommended improving AI design and training to boost workers' metacognitive
abilities for more effective collaboration.

@@

CharlieBot is an educational chatbot powered by ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) \citep{Sun24}. RAG is an AI technique that
integrates information retrieval with generative models. It first retrieves
relevant documents or data from a knowledge base or external source using a
retrieval model. Then, a generative model uses this retrieved information to
generate more accurate, context-aware responses. This dual-step process enhances
the quality and reliability of the chatbot's answers. A prior study on
CharlieBot's performance revealed that most students found its responses
appropriate for the Java programming course \citep{Hoyos24}.

@@

The study comprised three phases: data collection, categorization, and analysis.
During the data collection phase, students enrolled in a second-semester Java
course at the University Carlos III of Madrid (UC3M) were introduced to
CharlieBot and allowed to use it without following any prescribed educational
methodology. All data were collected anonymously to ensure that interactions
could not be traced back to individual students or linked to their academic
performance.

During the categorization phase, the students' prompts were initially classified
into eight distinct categories using the Claude.ai \citep{claude} tool.
Subsequently, the authors of this study manually reviewed these classifications
to ensure accuracy and alignment with the research objectives.

Initially, \cite{Ghimire24} proposed four categories: Debugging Help (DH),
Conceptual Question (CQ), Code Snippet (CS), and Complete Solution (CSO).
However, the data collected indicated the need for additional categories,
leading to the inclusion of four more: Multiple Questions (MQ), Student
Corrections (SC), Language Change (LC), and Uncategorized (U). Unlike study
\cite{Ghimire24}, which analyzed students' behavior using artificial
intelligence for specific programming tasks, our study allowed participants to
explore the chatbot as they preferred. This approach enabled the emergence of
additional categories of analysis, broadening the understanding of the tool's
uses and interactions. Table \ref{tab:categories} presents these categories,
their descriptions, and an example. Besides that, the topics covered in the
students' prompts were categorized based on the course issues, which include
(1) Java, (2) Object Orientation, (3) Testing, (4) Recursion, (5) Data
Structures, and (6) Sorting and Searching Algorithms.

The analysis phase involved using Python/Pandas scripts to extract information
from previously classified data. The data sets, and code used in this process
are available at https://github.com/rodrigoprestesmachado/uc3m.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.61]{img/figure1.png}
  \caption{Categorization of messages.}
  \label{fig:graph1}
\end{figure}

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.41]{img/figure2.png}
  \caption{Examples of conversations.}
  \label{fig:graph2}
\end{figure*}

@@

\begin{table*}[htbp]
  \caption{Average and Standard Deviation of Different Categories (in minutes)}
  \begin{center}
    \renewcommand{\arraystretch}{1.1} % Adjust the spacing between rows
    \begin{tabular}{p{4cm} p{3cm} p{3cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
      \hline
      Code Snippet & 1.34 & 1.95 \\
      Uncategorized & 1.70 & 3.59 \\
      Language Change & 2.23 & 3.71 \\
      Multiple Question & 2.47 & 5.66 \\
      \hdashline
      \textit{Overall} & \textit{3.78} & \textit{6.95} \\
      \hdashline
      Student Correction & 4.07 & 8.42 \\
      Debugging Help & 5.30 & 10.80 \\
      Conceptual Questions & 5.47 & 10.46 \\
      Complete Solution & 7.65 & 11.03 \\
      \hline
    \end{tabular}
    \label{tab:averages}
  \end{center}
\end{table*}

@@

Conversations without spacing comprise around 66\% of the total, with 34.2\%
showing a topic change, indicating interleaving (equivalent to 22.8\% of all
conversations). Qualitative analysis of these conversations without spacing
revealed that about 7\% of the conversations presented interleaving caused by
students' at least one request to solve quizzes. However, the speed with which
students interact with the chatbot to solve quizzes may suggest a more
superficial reflection on the answers received (see section 4.2.1). On the other
hand, 8.8\% of the conversations showed no spacing but included topic changes,
suggesting that students effectively might employ a metacognitive interleaving
strategy to distinguish concepts around a problem. Finally, 7\% of the
conversations featured interaction pauses lasting between 20 and 59 minutes but
fell short of the study's definition of spacing, which required pauses of over
60 minutes. This subset of conversations revealed a natural interleaving of
topics, reflecting a chatbot usage pattern.

@@

The analysis categories, spacing, and interleaving markers provided insights
into students' interactions with the chatbot. With the concept of spacing, it
was possible to define students' study sessions, which, in this work, consisted
of a few interaction messages with the chatbot. Consequently, a significant
portion of the observed interleaving occurred due to spacing, that is, in the
transition between study sessions. These findings indicate that the intentional
use of learning strategies, such as spacing and interleaving, is still limited.
Therefore, encouraging the deliberate application of these practices can enhance
students' learning outcomes using generative artificial intelligence (GenAI)
tools.

Messages from the Code Snippets and Multiple Question (Quiz) categories stood
out for involving consistently quick interactions with the bot. This pattern
suggests some level of engagement with the chatbot but also raises the
possibility of less in-depth reflection on the content received, especially when
students request small code snippets or the resolution of exercises.

@@

There are ways to deepen this study; based on previous references, 60 minutes
was the standard spacing measure. However, further studies will deepen this
analysis using a quantitative variable to better represent different spacing
intervals. Future research can also include additional metacognitive strategies,
which will involve interpreting user messages in depth. For instance, the
Monitoring strategy was evident when students expressed that a concept remained
unclear and requested further clarification. Another necessary analysis involves
investigating the relationship between student profiles, metacognitive
strategies for using GenAI, and learning outcomes.
